%. Pedigrees, allele frequencies, and lineages
% * recombination, segregation, and mutation
% * Moran and Cannings models
%   * including general, nonhomogeneous
%   * as generating a pedigree
% * forwards and reverse processes
% * limiting, continuous process


%%%%%%% %%%%%%%%
\section*{General goal}

We are presented with data: genome sequences plus other information,
about a number of individuals that are usually random samples from some population(s).
No matter what our analysis goal --
estimate levels of relatedness between the samples;
infer their ancestral genome sequence(s);
identify genomic locations subject to selection
-- we must understand the relationship between process parameters
(e.g.\ migration rates, selection coefficients)
and the observed genomic patterns of dissimilarity.



%%%%%%% %%%%%%%%
\section{Recombination, segregation, and mutation}

First, a reminder about how reproduction makes a new genome out of the genomes of the parents, in diploids.
There are exceptions (of course), but mostly,
each organism has two copies of each chromosome (counting sex chromosomes as a pair), one from mom and one from dad.
Offspring are the union of two gametes (an egg and a spermatid),
and each gamete is produced by a single diploid cell:
\begin{itemize}
    \item duplicating each chromosome (by mitosis)
    \item recombining the homlogous copies
    \item segregating these copies among four daughter cells.
\end{itemize}
Sometimes all four daughter cells become gametes;
sometimes (e.g.\ female meiosis in mammals) only one will.


\paragraph{Recombination} is complicated, varies along the genome, and is affected by genomic factors and motifs.
To make it tractable, we assume that recombination breakpoints occur as a Poisson point process 
with constant mean of 2 breakpoint per unit of length,
and that each breakpoint is a crossing over between a randomly chosen maternal and a randomly chosen paternal chromosome.
(This produces an average of 1 crossover per chromosome per unit length, 
i.e.\ we are measuring length in ``genetic map length''.)


\paragraph{Mutation}
Evolution needs ``descent \emph{with modification}'' -- so we need mutation, also.
At the level of detail we use here, we'll simply model mutation as another Poisson point process --
suppose that each gamete differs from its progenitor chromosome at the points of an Poisson point process;
usually we assume this has mean rate $\mu$ per unit of map length,
but in reality this rate varies along the genome.

\paragraph{Segregation}
Once four gametes are produced, it remains to be decided which of the four produce the offspring.
We assume that one is chosen uniformly at random,
independently of the result of recombination.
(Again, there are exceptions.)


%%%%%%% %%%%%%%%
\section{Mate choice: pedigrees}

Now that we have a model for how to produce offspring from two parents,
to model how genes move around in a population,
we need a model for how mates are chosen.
There are various models of ``random mating''.

\paragraph{The diploid Moran model:}
Continuous-time, fixed population size, overlapping generations.
In continuous time, each individual, at rate 1, chooses another uniformly at random
with whom to mate;
they produce one offspring
that replaces a randomly chosen individual (possibly including the parents).

\paragraph{The diploid Cannings model:}
Discrete time, varying population size, nonoverlapping generations.
Suppose at time $t$ there are $N_t$ members of the population,
which we take as a given trajectory.
Each pair of individuals could potentially produce some offspring;
let $X_{ij}$ denote the number produced in this event by pair $(i,j)$ with $i$ as the mother and $j$ as the father for each $1 \le i,j \le N_{t}$.
Suppose that the $X_{ij}$ are \emph{exchangeable},
i.e.\ that $(X_{ij})_{i,j=1}^{N_{t}} \deq (X_{\pi(i)\pi(j)})_{i,j=1}^{N_{t}}$ for any permutation $\pi$ of $(1,2,\ldots,N_{t})$,
and that $\sum_{ij} X_{ij} = N_{t+1}$.
Note that the organisms could be hermaphrodite or unisexual,
as long as we assume that sex determination is independent of siblingship,
and so effectively make sex determination the first step of reproduction.


Each of these produces a \emph{random pedigree}, 
i.e.\ a directed graph with nodes indexed by $(t,k)$, for $1 \le k \le N_t$
and two types of edges, corresponding to maternal versus paternal relationships.
Each arrow represents one meiosis, i.e.\ the result of recombination and segration to produce a gamete.
If we can assume that genetic material does not affect mate choice,
then a model for a population can first choose a random pedigree,
then determine genetic relatedness by making the choices of recombination and segregation
independently in each meiosis.



%%%%%%% %%%%%%%%
\section*{Aside: The genome is not passive}

Of course, the genes an individual carries can have quite a strong influence
on their mate choice, number of offspring, and even on the outcome of recombination and segregation.
In particular, once we know the genomes of every individual, the Cannings model makes no sense,
since individuals are clearly not exchangeable.
This makes things much more complicated,
and we do our best to ignore it,
for instance, by imagining we are tracking only segments of the genome not under selection,
so that genetic variation in fitness only contributes to the distribution of offspring number.


%%%%%%% %%%%%%%%
\section{Summary statistics}

Forget about the specifics of the mate choice models for the moment.
First let's think about the sorts of summary statistics we can compute from the data.
First, we fix some notation.
For a sample of individuals indexed by some set $A$,
genotyped at a set of genomic positions indexed by $S$,
the data are $\{G_{ijk} \; : \; i \in A \; j \in S \; k \in \{m,p\} \}$,
i.e.\ $G_{ijm}$ is allele that the $i^\mathrm{th}$ individual inherited at the $j^\mathrm{th}$ position from her mother,
and $G_{ijp}$ is the corresponding allele inherited from her father.


\paragraph{Heterozygosity} in a group of individuals in a genomic region 
is the probability that a randomly chosen individual is heterozygous at a randomly chosen nucleotide,
or 
\begin{align}
  H_O = \frac{ \# \left\{ (i,j) \st i \in A \; j \in S \; G_{ijp} \neq G_{ijm} \right\} }{ |S|\,|A| } ,
\end{align}
where $|S|$ denotes the total number of loci and $|A|$ denotes the total number of individuals.

In other words, $H_O$ is the proportion of homologous alleles that differ from each other, across $S$ and across $A$.
By calling them ``homologous'' we assume they share a common ancestor;
so if they differ there must have occurred a mutation since that common ancestor.
Take a single individual $i$,
suppose that the chance of a mutation occurring at site $j$ in a particular meiosis is $\mu$,
and that there have been $\tau_{ij}$ generations since the common ancestor of the maternal and paternal copies.
The probability that there has been no mutations since that time is $(1-\mu)^{2 \tau_{ij}}$,
since there are $2 \tau_{ij}$ meioses separating the two.
The proportion of heterozygous sites is determined by the empirical distribution 
of times back to the most common ancestor of paired homologous sites, averaged across sites and across individuals.
Let $\tau_H$ denote this distribution, i.e.\ $\P\{ \tau_H = t \} = \#\{ (i,j) \st \tau_{ij} = t \} / |S||A|$.
If we omit back mutation, so that, once mutated, the two can never be identical again,
then
\begin{align}
  H_O &= \frac{1}{|S|\,|A|} \sum_{i \in A} \sum_{j \in S} (1-\mu)^{2 \tau_{ij}} \\
    &= \E\left[ (1-\mu)^\tau_H \right] \\
    &\approx \E\left[ e^{-\mu \tau_H } \right] .
\end{align}
Note that $\tau_H$, if it was observable, would be a good \emph{summary statistic} (albeit complicated) of the pedigree,
and depends implicitly on the choice of individuals $A$ and the choice of genomic region $S$.
If $\mu$ is small, then $H_O \approx \mu \E[ \tau_H ]$,
i.e.\ the proportion of sites that an individual is heterozygous
is equal to the mutation rates multiplied by the average time back to the common ancestor of the maternal and paternal chomosomes.



\paragraph{Mean number of pairwise differences}, a.k.a.\ ``expected heterozygosity'':
A similar statistic is the chance that two randomly chosen alleles from $A$ at a random site in $S$ differ:
\begin{align}
  H_E &= \frac{ \#\{ (j,i_1,i_2,k_1,k_2) \st G_{i_1jk_1} \neq G_{i_2jk_2}  \} }{2|S|\,|A|(|A|-1)}  .
\end{align}
If, as above, we define $\tau$ to be the random number of generations back to the most recent common ancestor
between a uniformly chosen pair alleles in a random pair of individuals in $A$ at a randomly chosen site in $S$,
then again
\begin{align}
  H_E &= \E\left[ (1-\mu)^\tau ] \approx \E\left[ e^{-\mu \tau} \right] .
\end{align}




%%%%%%% %%%%%%%%
\section{Allele frequencies: forwards time}

The processes described above keep track of what every individuals' genome is at each point in time.
Now, we need to compute something.
To get our hands on something concrete, consider the marginal process at a single nonrecombining locus.
suppose that there are two possible variants at this locus,
label them `0' and `1',
and denote the frequency of `1's in the population at time $t$ by $P_t$.
Also neglect the influence of mutation.
The change in this frequency is equal to the difference in aggregate numbers offsprings produced by individuals 
carrying each of the two variants,
divided by the population size.
If the locus is neutral
-- i.e.\ an organism's mate choice and offspring numbers are independent of the alleles carried --
then the average number of 

\paragraph{Diploid Moran model:}


%%%%%%% %%%%%%%%
\section{Haploid models}

Take each chromosome as an individual\ldots

\paragraph{The haploid Moran model:}
Continuous-time, fixed population size.
In continuous time, each individual, at rate 1, 
produces one offspring (through simple division)
that replaces a randomly chosen individual (possibly the parent).

\paragraph{The haploid Wright--Fisher model:}
Discrete-time, varying population size.
Suppose at time $t$ there is room for $N_t$ members of the population,
which we take as a given trajectory.
Each individual produces a Poisson number of offspring with large mean;
of the total pool of offspring, a uniformly chosen set of $N_{t+1}$ of these
are chosen to form the next generation.
